---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# dosefitr

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/license/mit)
<!-- badges: end -->

End-to-end dose-response analysis for **NanoBRET kinase binding assays** and
**cell viability experiments**. `dosefitr` reads raw BMG PHERAstar plate-reader
exports, computes BRET ratios (or processes luminescence), detects outliers
with the ROUT method, fits 3PL / 4PL logistic models, reports IC50 / pIC50 with
confidence intervals, and writes its results to Excel automatically — per-plate
result files plus a consolidated report — alongside publication-ready plots
(with optional, lab-specific Scarab-format submission tables).

## Hero example

```{r hero, eval = FALSE}
library(dosefitr)

# Stage the two bundled NanoBRET plates into a working directory
extdata_dir <- system.file("extdata", package = "dosefitr")
work_dir    <- file.path(tempdir(), "dosefitr_demo")
dir.create(work_dir, showWarnings = FALSE)
file.copy(
    list.files(extdata_dir, pattern = "^nanobret_", full.names = TRUE),
    work_dir
)

# 1. Compute BRET ratios across both plates
ratio_res <- batch_ratio_analysis(
    directory        = work_dir,
    info_file        = "nanobret_info.xlsx",
    data_pattern     = "nanobret_plate_\\d+\\.xlsx$",
    control_0perc    = "1",
    control_100perc  = "24",
    selected_columns = 1:24,
    generate_reports = FALSE,
    output_dir       = file.path(tempdir(), "dosefitr_ratio")
)

# 2. Fit 4PL curves and 3. save panel plots
drc_res <- batch_drc_analysis(ratio_res, model = "4pl",
                              generate_reports = FALSE,
                              output_dir = file.path(tempdir(), "dosefitr_drc"))
batch_save_all_drc_plots(drc_res,
                         output_dir = file.path(tempdir(), "dosefitr_plots"))
```

## Installation

`dosefitr` is planned for submission to Bioconductor once the accompanying
manuscript is published. Once accepted, install the release version with:

```{r install-bioc, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dosefitr")
```

The in-development version is on GitHub:

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("AKKDataAnalysis/dosefitr")
```

System requirements: R (>= 4.4.0). All package dependencies (including
`OptimModel`, which provides the ROUT outlier test) are on CRAN and
`remotes::install_github()` pulls them automatically.

## What it does

`dosefitr` turns raw 384-well plate-reader exports into a full dose-response
analysis package in one R session, without stitching together plate-parsing,
curve-fitting, and export tools:

- **NanoBRET pipeline** (dual-channel donor/acceptor): `batch_ratio_analysis`
  reads BMG PHERAstar exports, extracts donor and acceptor matrices, computes
  BRET ratios per well, and processes each row against its DMSO / saturated
  controls, writing a per-plate result workbook plus a consolidated report.
- **Viability pipeline** (single-channel luminescence, e.g. CellTiter-Glo):
  `batch_viability_analysis` reads single-channel exports, processes against DMSO,
  and clamps the 0 % floor at the compound-saturated wells, writing per-plate
  result workbooks plus quality-metric files. Three info-table /
  control layouts are supported via the `version = "v1"` / `"v2"` / `"v3"`
  switch: **v1** (default) expects two technical replicates per plate row and
  applies per-construct / per-row / global control means; **v2** is the
  NanoBRET-style processor, using a fixed 0 % scalar and one or more plate
  columns averaged for the 100 % control; **v3** encodes one replicate per row
  (replicate rows share `Target` + `Compound` in the info table).
- **Shared downstream stack**: `rout_outliers_batch` flags outlier wells with
  the ROUT method (Motulsky & Brown 2006); `batch_drc_analysis` fits 3PL or 4PL
  models with assay-aware plausibility limits on Bottom, Top, and Hill slope;
  `batch_save_all_drc_plots` writes one plot per compound plus a
  patchwork-assembled per-plate panel; `plot_multiple_compounds` overlays
  compounds in one figure; `compare_plates_drc` compares the same compound
  across plates.
- **Automatic QC**: Z-prime, assay window, signal-to-background, and per-plate
  CV%, plus a `Curve_Quality` label per compound
  (`Good curve`, `Wide logIC50 CI range`, `Top too low`, ...) exposed on the
  per-plate summary table.
- **Export**: result export is built into the batch functions themselves —
  `batch_ratio_analysis`, `batch_viability_analysis`, and `batch_drc_analysis`
  each write their results to Excel (per-plate files plus a consolidated report)
  whenever `generate_reports = TRUE` (the default), so a standard run produces its
  result workbooks with no extra step. `save_multiple_sheets` writes any number of
  data frames to a single workbook. For laboratories that submit to the SGC
  Scarab system, the optional `scarab_table` (NanoBRET) and `scarab_viability`
  helpers additionally emit lab-specific Scarab-format submission tables.

## What it doesn't do

`dosefitr` does not read raw fluorescence images, does not perform
competition-binding thermodynamics beyond IC50 / pIC50, and is not designed for
combination-drug (Bliss / Loewe / HSA) analysis. For large-scale
pharmacogenomic meta-analyses across cell-line panels and public screens,
see `PharmacoGx` on Bioconductor.

## Pipeline overview

```
Raw Excel files
      |
      v
batch_ratio_analysis()        <- NanoBRET:   reads plates, computes BRET ratios
batch_viability_analysis()    <- Viability:  reads plates, processes signal
      |
      v
rout_outliers_batch()         <- detects and removes outliers (optional)
      |
      v
batch_drc_analysis()          <- fits dose-response curves (3PL or 4PL)
      |
      |--> batch_save_all_drc_plots() <- one plot per compound + panel per plate
      |--> plot_multiple_compounds()  <- overlays selected compounds
      |--> compare_plates_drc()       <- same compound across plates
      |--> scarab_table()             <- optional Scarab export, lab-specific (NanoBRET)
      `--> scarab_viability()         <- optional Scarab export, lab-specific (viability)
```

## Function map

| Group                | Functions                                                                                              |
|----------------------|--------------------------------------------------------------------------------------------------------|
| Reading / processing| `batch_ratio_analysis` `batch_viability_analysis` `process_viability_data` `process_viability_data_v2` `process_viability_data_v3` `ratio_dose_response` `ratio_dose_response_v2` `merge_plate_replicates` |
| Import (pre-computed)| `batch_read_tables`                                                                                    |
| Quality control      | `rout_outliers` `rout_outliers_batch`                                                                  |
| Fitting              | `batch_drc_analysis` `fit_drc_4pl` `fit_drc_3pl` `reshape_dr_table`                                    |
| Plotting             | `batch_save_all_drc_plots` `plot_multiple_compounds` `plot_dose_response` `plot_all_dose_responses` `compare_plates_drc` `plot_outliers_curves` `plot_outliers_batch_curves` `plot_drc_batch` |
| Export               | `save_multiple_sheets` `scarab_table` `scarab_viability` (batch functions also write result workbooks directly) |

## Documentation

Four vignettes ship inside the package. After installing, browse the full
list in your R session with:

```{r browse-vignettes, eval = FALSE}
browseVignettes("dosefitr")
```

Or open each one directly:

- **`vignette("dosefitr")`** — introduction and package tour: motivation,
  comparison with related dose-response packages, and pointers into the
  pipeline vignettes.
- **`vignette("dosefitr-nanobret")`** — end-to-end NanoBRET walkthrough on the
  bundled example plates (BRET ratios, ROUT, 3PL / 4PL fits, plotting, optional
  Scarab export).
- **`vignette("dosefitr-viability")`** — end-to-end cell-viability walkthrough
  on the bundled example plates (single-channel processing, 3PL / 4PL fits,
  optional Scarab export).
- **`vignette("dosefitr-protocol")`** — the day-to-day analysis script,
  section by section, mirroring how you would run it against your own plates
  (ratios, ROUT, replicate merging, DRC fits, plotting, optional Scarab metadata).

The vignettes above are the primary, executable tutorials. As a convenience,
the same day-to-day workflows are also shipped as standalone, copy-paste R
scripts you can adapt to your own experiment (edit the paths / control
columns and run top to bottom):

- `Example_protocol_NanoBRET.R` — NanoBRET BRET-ratio workflow.
- `Example_protocol_Viability.R` — cell-viability workflow.
- `Example_protocol_analysis.R` — combined analysis walkthrough.

Locate them in your installed copy with:

```{r examples-path, eval = FALSE}
system.file("examples", package = "dosefitr")
list.files(system.file("examples", package = "dosefitr"), pattern = "\\.R$")
```

or browse them on GitHub under
[`inst/examples/`](https://github.com/AKKDataAnalysis/dosefitr/tree/main/inst/examples).
These scripts are a lightweight complement to the vignettes, not a replacement
— the vignettes remain the authoritative, reproducible documentation.

Each exported function also has its own help page; open the grouped index
with `help(package = "dosefitr")`.

## Citation

Manuscript in preparation. Run `citation("dosefitr")` for the up-to-date
record. The ROUT outlier-detection algorithm used by `rout_outliers()` was
developed by Motulsky and Brown (2006); please cite the original paper:

> Motulsky, H. J., & Brown, R. E. (2006). Detecting outliers when fitting
> data with nonlinear regression -- a new method based on robust nonlinear
> regression and the false discovery rate. *BMC Bioinformatics*, 7, 123.
> [doi:10.1186/1471-2105-7-123](https://doi.org/10.1186/1471-2105-7-123)

## Getting help

- Function-level docs: `help(package = "dosefitr")`
- Bug reports and feature requests:
  <https://github.com/AKKDataAnalysis/dosefitr/issues>

## License

MIT (c) 2025 Thiago Loreto Matos. See `LICENSE` for details.
