---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.path  = "man/figures/README-",
  out.width = "100%"
)
```

# dosefitr

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/license/mit)
<!-- badges: end -->

End-to-end dose-response analysis for **NanoBRET kinase binding assays** and
**cell viability experiments**. `dosefitr` reads raw BMG PHERAstar plate-reader
exports, computes BRET ratios (or normalises luminescence), detects outliers
with the ROUT method, fits 3PL / 4PL logistic models, reports IC50 / pIC50 with
confidence intervals, and exports publication-ready plots, multi-sheet Excel
workbooks, and Scarab-formatted submission tables.

## Hero example

```{r hero, eval = FALSE}
library(dosefitr)

# Stage the two bundled NanoBRET plates into a working directory
extdata_dir <- system.file("extdata", package = "dosefitr")
work_dir    <- file.path(tempdir(), "dosefitr_demo")
dir.create(work_dir, showWarnings = FALSE)
file.copy(
    list.files(extdata_dir, pattern = "^nanobret_", full.names = TRUE),
    work_dir
)

# 1. Compute BRET ratios across both plates
ratio_res <- batch_ratio_analysis(
    directory        = work_dir,
    info_file        = "nanobret_info.xlsx",
    data_pattern     = "nanobret_plate_\\d+\\.xlsx$",
    control_0perc    = "1",
    control_100perc  = "24",
    selected_columns = 1:24,
    generate_reports = FALSE,
    output_dir       = file.path(tempdir(), "dosefitr_ratio")
)

# 2. Fit 4PL curves and 3. save panel plots
drc_res <- batch_drc_analysis(ratio_res, model = "4pl",
                              generate_reports = FALSE,
                              output_dir = file.path(tempdir(), "dosefitr_drc"))
batch_save_all_drc_plots(drc_res,
                         output_dir = file.path(tempdir(), "dosefitr_plots"))
```

## Installation

`dosefitr` is planned for submission to Bioconductor once the accompanying
manuscript is published. Once accepted, install the release version with:

```{r install-bioc, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dosefitr")
```

The in-development version is on GitHub:

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("AKKDataAnalysis/dosefitr")
```

System requirements: R (>= 4.4.0). All package dependencies (including
`OptimModel`, which provides the ROUT outlier test) are on CRAN and
`remotes::install_github()` pulls them automatically.

## What it does

`dosefitr` turns raw 384-well plate-reader exports into a full dose-response
analysis package in one R session, without stitching together plate-parsing,
curve-fitting, and export tools:

- **NanoBRET pipeline** (dual-channel donor/acceptor): `batch_ratio_analysis`
  reads BMG PHERAstar exports, extracts donor and acceptor matrices, computes
  BRET ratios per well, and normalises each row by its DMSO / saturated
  controls.
- **Viability pipeline** (single-channel luminescence, e.g. CellTiter-Glo):
  `batch_viability_analysis` reads single-channel exports, normalises to DMSO,
  and clamps the 0 % floor at the compound-saturated wells. Two info-table
  layouts are supported via the `version = "v1"` / `"v2"` switch.
- **Shared downstream stack**: `rout_outliers_batch` flags outlier wells with
  the ROUT method (Motulsky & Brown 2006); `batch_drc_analysis` fits 3PL or 4PL
  models with assay-aware plausibility limits on Bottom, Top, and Hill slope;
  `batch_save_all_drc_plots` writes one plot per compound plus a
  patchwork-assembled per-plate panel; `plot_multiple_compounds` overlays
  compounds in one figure; `compare_plates_drc` compares the same compound
  across plates.
- **Automatic QC**: Z-prime, assay window, signal-to-background, and per-plate
  CV%, plus a `Curve_Quality` label per compound
  (`Good curve`, `Wide logIC50 CI range`, `Top too low`, ...) exposed on the
  per-plate summary table.
- **Export**: `scarab_table` (NanoBRET) and `scarab_viability` produce
  Scarab-format submission tables for SGC-style kinase profiling submission;
  `save_multiple_sheets` writes any number of data frames to one Excel file.

## What it doesn't do

`dosefitr` does not read raw fluorescence images, does not perform
competition-binding thermodynamics beyond IC50 / pIC50, and is not designed for
combination-drug (Bliss / Loewe / HSA) analysis. For large-scale
pharmacogenomic meta-analyses across cell-line panels and public screens,
see `PharmacoGx` on Bioconductor.

## Pipeline overview

```
Raw Excel files
      |
      v
batch_ratio_analysis()        <- NanoBRET:   reads plates, computes BRET ratios
batch_viability_analysis()    <- Viability:  reads plates, normalises signal
      |
      v
rout_outliers_batch()         <- detects and removes outliers (optional)
      |
      v
batch_drc_analysis()          <- fits dose-response curves (3PL or 4PL)
      |
      |--> batch_save_all_drc_plots() <- one plot per compound + panel per plate
      |--> plot_multiple_compounds()  <- overlays selected compounds
      |--> compare_plates_drc()       <- same compound across plates
      |--> scarab_table()             <- Scarab-format export (NanoBRET)
      `--> scarab_viability()         <- Scarab-format export (viability)
```

## Function map

| Group                | Functions                                                                                              |
|----------------------|--------------------------------------------------------------------------------------------------------|
| Reading / normalising| `batch_ratio_analysis` `batch_viability_analysis` `process_viability_data` `process_viability_data_v2` `ratio_dose_response` `ratio_dose_response_v2` `merge_plate_replicates` |
| Quality control      | `rout_outliers` `rout_outliers_batch`                                                                  |
| Fitting              | `batch_drc_analysis` `fit_drc_4pl` `fit_drc_3pl` `reshape_dr_table`                                    |
| Plotting             | `batch_save_all_drc_plots` `plot_multiple_compounds` `plot_dose_response` `plot_all_dose_responses` `compare_plates_drc` `plot_outliers_curves` `plot_outliers_batch_curves` `plot_drc_batch` |
| Export               | `scarab_table` `scarab_viability` `save_multiple_sheets`                                               |

## Documentation

Three vignettes ship with the package:

- **`vignette("dosefitr")`** — introduction, motivation, and comparison with
  related dose-response packages.
- **`vignette("dosefitr-nanobret")`** — end-to-end NanoBRET walkthrough on the
  bundled example plates (BRET ratios, ROUT, 4PL fits, plotting, Scarab
  export).
- **`vignette("dosefitr-viability")`** — end-to-end viability walkthrough on
  the bundled example plates (single-channel normalisation, 4PL fits, Scarab
  export).

Each exported function has its own help page; open the grouped index with
`help(package = "dosefitr")`.

## Citation

Manuscript in preparation. Run `citation("dosefitr")` for the up-to-date
record. The ROUT outlier-detection algorithm used by `rout_outliers()` was
developed by Motulsky and Brown (2006); please cite the original paper:

> Motulsky, H. J., & Brown, R. E. (2006). Detecting outliers when fitting
> data with nonlinear regression -- a new method based on robust nonlinear
> regression and the false discovery rate. *BMC Bioinformatics*, 7, 123.
> [doi:10.1186/1471-2105-7-123](https://doi.org/10.1186/1471-2105-7-123)

## Getting help

- Function-level docs: `help(package = "dosefitr")`
- Bug reports and feature requests:
  <https://github.com/AKKDataAnalysis/dosefitr/issues>

## License

MIT (c) 2025 Thiago Loreto Matos. See `LICENSE` for details.
