---
title: "End-to-end NanoBRET analysis protocol"
author:
  - name: Thiago Loreto Matos
    affiliation: |
      Pharmacogenetics Laboratory, Drug Research and Development Center,
      Department of Physiology and Pharmacology, Federal University of Ceará,
      Fortaleza, 60430-160, Brazil
    email: thiago-loreto@hotmail.com
date: "`r Sys.Date()`"
package: dosefitr
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{End-to-end NanoBRET analysis protocol}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  eval       = TRUE,
  message    = FALSE,
  warning    = FALSE
)
```

```{=html}
<style>
p.author-name       { font-size: 1.05em; margin-top: -0.5em; font-style: italic; font-weight: 700; }
p.author-affiliation{ font-size: 0.88em; color: #444; margin-top: 0.2em; font-style: normal; font-weight: normal; }
p.author-email      { font-size: 0.88em; margin-top: 0.2em; font-style: normal; font-weight: normal; }
p.author-affiliation .affil-mark,
p.author-email .affil-mark { font-style: normal; font-weight: normal; }

h4.author           { font-size: 1.05em; margin: 0.9em 0 0.05em 0; font-style: italic; font-weight: 700; }
address.author_afil { display: block; font-style: normal; margin: 0 0 0.1em 0; }
address.author_afil p { font-size: 0.88em; color: #444; margin: 0 0 0.05em 0; }
address.author_afil br { display: none; }
a.author_email[href="mailto:#"] { display: none; }
address.author_afil a.email { font-size: 0.88em; }
h4.date             { font-size: 0.95em; margin-top: 1em; font-weight: 500; }
</style>
```

# Introduction

This vignette is an end-to-end **NanoBRET analysis protocol**, written as
a working template you can adapt to your own experiments. Each stage of
the analysis is a separate topic: one short prose block explaining what
the step does and how to change it for your project, followed by a code
block that runs against the two bundled example plates.

The walkthrough mirrors the pattern of a real-world `dosefitr` session:
locate the raw data, build the ratio table, detect outliers, fit the
dose-response curves, save the plots, compare across plates, and export a
Scarab-formatted submission table. Where a step has a common variation
(v2 ratio processor, viability pipeline, replicate merging), that
variation is treated as its own sub-topic instead of being folded into
the main flow.

For an overview of the package and a comparison with other CRAN and
Bioconductor dose-response packages, see
`vignette("dosefitr", package = "dosefitr")`. For a per-function
walkthrough of the NanoBRET pipeline, see
`vignette("dosefitr-nanobret", package = "dosefitr")`. For the
single-channel viability pipeline, see
`vignette("dosefitr-viability", package = "dosefitr")`.

# Install and load

Install the release version from Bioconductor once accepted:

```{r install-bioc, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dosefitr")
```

Or install the in-development version from GitHub:

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("AKKDataAnalysis/dosefitr")
```

Load the package:

```{r libs}
library(dosefitr)
```

# Point at your data

`batch_ratio_analysis()` scans a directory for plate files that match a
regular expression, together with a single Excel info table that maps
each plate-row to a compound. Everything downstream builds on the object
that this function returns.

The package ships two example NanoBRET plates plus an info table under
`inst/extdata/`. To keep this vignette compliant with the "no writes
outside `tempdir()`" rule, we stage those files into a working directory
under `tempdir()`:

```{r stage}
extdata_dir <- system.file("extdata", package = "dosefitr")
work_dir    <- file.path(tempdir(), "dosefitr_protocol")
dir.create(work_dir, showWarnings = FALSE, recursive = TRUE)

invisible(file.copy(
    from      = list.files(extdata_dir, pattern = "^nanobret_",
                           full.names = TRUE),
    to        = work_dir,
    overwrite = TRUE
))

list.files(work_dir)
```

**In your own project.** Skip the copy-to-tempdir step and pass the
path to your data folder as the `directory` argument to
`batch_ratio_analysis()` below (for example,
`directory = "C:/path/to/your/dataset"`).

# Compute BRET ratios (v1)

`batch_ratio_analysis()` reads each plate's donor and acceptor channels,
computes the BRET ratio per well as `(acceptor / donor) * 1000`, and
normalises each compound row by the plate's 0 % and 100 % controls. The
result is a named list, one entry per plate, keyed by the info-table
sheet name.

The example call below uses the bundled data's layout (controls at
columns 1 and 24) and passes `low_value_threshold = 3000`, which flags
wells whose donor channel falls below 3000 counts as unreliable. On the
bundled data no wells fall below that floor, so no wells are flagged --
but the argument is included to demonstrate the pattern.

```{r ratio}
ratio_out <- file.path(tempdir(), "dosefitr_protocol_ratio")
dir.create(ratio_out, showWarnings = FALSE)

ratio_res <- batch_ratio_analysis(
    directory           = work_dir,
    info_file           = "nanobret_info.xlsx",
    data_pattern        = "nanobret_plate_\\d+\\.xlsx$",
    control_0perc       = "1",
    control_100perc     = "24",
    selected_columns    = 1:24,
    low_value_threshold = 3000,
    generate_reports    = FALSE,
    output_dir          = ratio_out,
    verbose             = FALSE
)

names(ratio_res)

mrt <- ratio_res$plate_01$result$modified_ratio_table
dim(mrt)
head(mrt[, 1:5], 6)
```

**In your own project.** The three arguments most likely to differ
between plates are:

- `control_0perc` -- which column holds the 0 % control (baseline).
- `control_100perc` -- which column holds the 100 % control (saturated).
- `selected_columns` -- which columns to include in the analysis.

For example, if your DMSO baseline is in column 24 and your saturated
compound is in column 12:

```{r own-project-ratio, eval = FALSE}
ratio_res <- batch_ratio_analysis(
    directory           = "C:/path/to/your/dataset",
    control_0perc       = 24,
    control_100perc     = 12,
    low_value_threshold = 3000,
    output_dir          = "./drc_quality",
    verbose             = TRUE
)
```

# Compute BRET ratios (v2 processor)

The v2 processor handles two layouts the v1 processor does not:

1. **Info tables with merged control cells** -- the info sheet has a
   merged control block instead of one row per control column.
2. **Multiple 100 % control columns** -- `control_100perc` is a *vector*
   of column indices (for example, when the saturated-compound wells are
   duplicated across two columns).

Switch on the v2 processor by passing `function_version = "v2"`. The
snippet below shows the pattern; it does not run because the bundled
info table follows the v1 layout:

```{r ratio-v2, eval = FALSE}
ratio_res_v2 <- batch_ratio_analysis(
    directory           = "C:/path/to/your/dataset",
    control_0perc       = 16,
    control_100perc     = c(12, 24),
    low_value_threshold = 3000,
    function_version    = "v2",
    output_dir          = "./drc_quality",
    verbose             = TRUE
)
```

See `vignette("dosefitr-viability", package = "dosefitr")` for the
canonical v2-processor example on cell-viability data.

# Detect outliers (ROUT)

`rout_outliers_batch()` runs the Motulsky & Brown ROUT test
(BMC Bioinformatics, 2006) on every fitted curve. Wells that diverge by
more than the ROUT `Q` threshold **and** disagree with their replicate
are flagged.

You have a choice at this step: feed the raw `ratio_res` into the DRC
fitter and let the fitter see every point, or feed the ROUT-cleaned
`rout_res` and drop the outliers before fitting. Both are valid; the
outlier report is useful either way for QC.

```{r rout}
rout_res <- rout_outliers_batch(ratio_res, Q = 0.01, verbose = FALSE)

nrow(rout_res$outlier_summary)
head(rout_res$outlier_summary, 5)
```

Overlay the flagged wells on the fitted curves for a visual sanity check:

```{r rout-plot, fig.width = 8, fig.height = 6}
pobc_out <- file.path(tempdir(), "dosefitr_protocol_pobc")
dir.create(pobc_out, showWarnings = FALSE)

pobc_res <- plot_outliers_batch_curves(
    rout_res,
    plates     = "plate_01",
    output_dir = pobc_out,
    verbose    = FALSE
)
```

# Merge replicate plates (optional)

If the same 384-well layout was run on separate days or with different
compound batches, `merge_plate_replicates()` combines the per-plate
tables into a single wide table with all replicate columns side-by-side.
Bridging compounds (present on both plates) end up with four replicate
columns instead of two.

```{r merge}
merge_out <- file.path(tempdir(), "dosefitr_protocol_merge")
dir.create(merge_out, showWarnings = FALSE)

merged_res <- merge_plate_replicates(
    ratio_res,
    output_dir       = merge_out,
    generate_reports = FALSE
)

names(merged_res)

merged_mrt <- merged_res[[1]]$result$modified_ratio_table
dim(merged_mrt)
head(colnames(merged_mrt), 8)
```

Compounds that appear on only one plate keep their two-replicate columns;
compounds that appear on both plates gain four columns
(`Cpd1`, `Cpd1.2`, `Cpd1.3`, `Cpd1.4`).

# Fit dose-response curves (4PL / 3PL)

`batch_drc_analysis()` fits every compound with a 3- or 4-parameter
logistic model, using the `OptimModel`-based solver with assay-aware
plausibility limits on Bottom, Top, and Hill-slope. Each fit is labelled
with a `Curve_Quality` tag (`Good curve`, `Wide logIC50 CI range`,
`Top too low`, `Bottom too high`, ...) that summarises the fit's health.

## `model = "3pl"` or `"4pl"`?

The 3PL model fixes the Hill slope to 1 and estimates only Bottom, Top,
and LogIC50. The 4PL model additionally estimates the Hill slope. Use
3PL when you want a standard, comparable IC50 across many compounds and
you have no strong prior evidence for cooperative binding; use 4PL when
the slope itself is a parameter you care about, or when several
compounds visibly show slopes far from 1.

## `normalize = TRUE` or `normalize = FALSE`?

For **NanoBRET** you can fit either the raw control-normalised ratio
table (`normalize = FALSE`) or re-scale the fit output to a 0 - 100 %
inhibition axis inside the fitter (`normalize = TRUE`). Both give the
same `LogIC50` -- the difference is only whether the reported Bottom and
Top are in ratio units or in percent. `normalize = TRUE` is the more
Prism-like presentation; `normalize = FALSE` keeps the numbers on the
original BRET-ratio scale.

For **viability** the recommended default is `normalize = TRUE`, which
clamps the fit to the 0 - 100 % viability window; see
`vignette("dosefitr-viability", package = "dosefitr")`.

```{r drc}
drc_out <- file.path(tempdir(), "dosefitr_protocol_drc")
dir.create(drc_out, showWarnings = FALSE)

drc_res <- batch_drc_analysis(
    batch_results    = rout_res,
    model            = "3pl",
    normalize        = TRUE,
    output_dir       = drc_out,
    generate_reports = FALSE,
    verbose          = FALSE
)

sum_01 <- drc_res$drc_results$plate_01$drc_result$summary_table
sum_01[, c("Compound", "LogIC50", "R_squared", "Curve_Quality")]
```

**In your own project.** Switch `model` to `"4pl"` if you want a
free-Hill fit. Set `normalize = FALSE` if you would rather see the fit
on the raw BRET-ratio scale. If you skipped the ROUT step above, pass
`ratio_res` directly instead of `rout_res`.

# Save all dose-response plots

`batch_save_all_drc_plots()` writes one PNG per compound plus (optionally)
a combined patchwork panel per plate. Filenames encode the plate and
`construct:compound` key, so a plate of 16 compounds becomes 16 PNGs (or
32 across two plates).

```{r save-plots}
plots_out <- file.path(tempdir(), "dosefitr_protocol_plots")
dir.create(plots_out, showWarnings = FALSE)

save_res <- batch_save_all_drc_plots(
    batch_drc_results = drc_res,
    output_dir        = plots_out,
    dpi               = 72,
    save_panel        = FALSE,
    verbose           = FALSE
)

length(list.files(plots_out, recursive = TRUE, pattern = "\\.png$"))
```

**In your own project.** For large batches, narrow the output with
`plates_to_plot` (character vector of plate names) or `compounds_to_plot`
(character vector of compound suffixes -- the part after `:` in the
compound key). Set `save_panel = TRUE` to also emit a combined
per-plate panel figure.

# Overlay multiple compounds

`plot_multiple_compounds()` overlays several curves on a single set of
axes. This is useful for target-specific SAR comparisons -- for example,
the six KinaseA compounds side-by-side:

```{r overlay, fig.width = 7, fig.height = 5}
overlay <- plot_multiple_compounds(
    drc_res,
    compound_indices  = 1:6,
    plate             = "plate_01",
    color_palette     = "colorblind",
    plot_title        = "",
    legend_text_size  = 12,
    axis_text_size    = 11,
    axis_title_size   = 12,
    save_plot         = NULL,
    verbose           = FALSE
)

overlay
```

**In your own project.** Replace `compound_indices` with the row indices
you want to overlay (the same order as in `drc_res$..$summary_table`).
Pass `save_plot = TRUE` to auto-save the plot under the plate's DRC output
directory, or a file path to save with an explicit name.

# Compare compounds across plates

`compare_plates_drc()` matches compounds across plates (via
`compare_by = "compound"`) or targets across plates
(via `compare_by = "construct"`) and produces side-by-side fits for each
match. It is the standard cross-plate consistency check.

The two bundled plates share four "bridging" compounds
(`KinaseA:Cpd1`, `KinaseA:Cpd5`, `KinaseB:Cpd9`, `KinaseB:Cpd13`). We
compare a two-compound subset to keep the output small:

```{r compare}
cmp_out <- file.path(tempdir(), "dosefitr_protocol_cmp")
dir.create(cmp_out, showWarnings = FALSE)

cmp_res <- compare_plates_drc(
    drc_res,
    compare_by        = "compound",
    selected_entities = c("KinaseA:Cpd1", "KinaseB:Cpd9"),
    min_plates        = 2,
    output_dir        = cmp_out,
    plot_dpi          = 72,
    verbose           = FALSE
)

length(cmp_res)
names(cmp_res)
```

**In your own project.** Drop `selected_entities` to compare every
bridging compound. Use `compare_by = "construct"` when you want the
same target aggregated across plates (useful for tracking compound
libraries against a single kinase across weeks of screening).

# Build the Scarab table

`scarab_table()` produces a Scarab-format submission table: one long
sheet with fitted parameters, quality metrics, and experiment metadata
side-by-side. The metadata block encodes information a curator needs to
reproduce the analysis (date, experimenter, tracer identity and
concentration, nLuc orientation, decimal-separator convention).

The values below are illustrative -- supply your own for each experiment.

```{r scarab}
scarab <- scarab_table(
    ratio_res, drc_res,
    plate_name                 = "plate_01",
    date                       = "240101",
    experimenter_abbrev        = "ME",
    nLuc_orientation           = "C",
    tracer_kd_app              = -7.5,
    tracer_concentration_used  = -7.5,
    tracer                     = "Tracer K10",
    decimal_separator          = ","
)

dim(scarab)
head(scarab[, 1:5], 6)
```

**In your own project.** Set `decimal_separator = "."` if your Scarab
instance expects US-style decimals. Set `nLuc_orientation` to `"N"` for
N-terminal fusion constructs. `date` is a free-form character string
using the convention your team already uses (here `YYMMDD`).

# Cell viability sidebar

The same plate layout can carry a single-channel viability readout
(e.g. CellTiter-Glo). Substitute `batch_viability_analysis()` for
`batch_ratio_analysis()` and pass `normalize = TRUE` to the DRC step:

```{r viability-sidebar, eval = FALSE}
via_res <- batch_viability_analysis(
    directory        = "C:/path/to/your/dataset",
    control_0perc    = 13,
    control_100perc  = 12,
    selected_columns = c(2:23),
    output_dir       = "./drc_quality",
    verbose          = TRUE
)

drc_res_via <- batch_drc_analysis(
    batch_results = via_res,
    model         = "3pl",
    normalize     = TRUE,
    output_dir    = "./drc_results",
    verbose       = TRUE
)
```

For a runnable end-to-end viability example, see
`vignette("dosefitr-viability", package = "dosefitr")`.

# Sanity check

For the bundled plates you should see 16/16 successful fits per plate and
mostly `Good curve` labels. If your own analysis shows many failures,
first check that `control_0perc` and `control_100perc` match your actual
plate layout -- a mis-specified control column is the single most common
source of failed fits.

# Session info

```{r session-info, echo = FALSE}
sessionInfo()
```
