---
title: "dosefitr walkthrough"
author:
  - name: Thiago Loreto Matos
    affiliation: |
      Pharmacogenetics Laboratory, Drug Research and Development Center,
      Department of Physiology and Pharmacology, Federal University of Ceará,
      Fortaleza, 60430-160, Brazil
    email: thiago-loreto@hotmail.com
date: "`r Sys.Date()`"
package: dosefitr
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{dosefitr walkthrough}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse  = TRUE,
    comment   = "#>",
    fig.align = "center"
)
```

```{=html}
<style>
/* Typography tuning for BiocStyle's semantic author block.
   Selectors match p.author-name, p.author-affiliation, p.author-email
   produced by BiocStyle::html_document (verified against rendered DOM). */
p.author-name       { font-size: 1.05em; margin-top: -0.5em; font-style: italic; font-weight: 700; }
p.author-affiliation{ font-size: 0.88em; color: #444; margin-top: 0.2em; font-style: normal; font-weight: normal; }
p.author-email      { font-size: 0.88em; margin-top: 0.2em; font-style: normal; font-weight: normal; }
p.author-affiliation .affil-mark,
p.author-email .affil-mark { font-style: normal; font-weight: normal; }

/* Pkgdown produces different author-block markup than BiocStyle
   (h4.author + address.author_afil + <a class=email>).
   These rules tighten the pkgdown site's author block only; they have
   no effect on BiocStyle's own vignette render because the selectors
   don't match BiocStyle's DOM. */
h4.author           { font-size: 1.05em; margin: 0.9em 0 0.05em 0; font-style: italic; font-weight: 700; }
address.author_afil { display: block; font-style: normal; margin: 0 0 0.1em 0; }
address.author_afil p { font-size: 0.88em; color: #444; margin: 0 0 0.05em 0; }
address.author_afil br { display: none; }
a.author_email[href="mailto:#"] { display: none; }
address.author_afil a.email { font-size: 0.88em; }
h4.date             { font-size: 0.95em; margin-top: 1em; font-weight: 500; }
</style>
```

# Introduction

`dosefitr` is a self-contained R toolchain for **NanoBRET kinase binding
assays** and **cell viability** experiments read out on the BMG PHERAstar
platform. It turns raw 384-well plate-reader exports into a full
dose-response analysis package in one R session: plate parsing, control
normalisation, ROUT outlier detection, 3PL / 4PL logistic fitting with
IC50 / pIC50 and confidence intervals, publication-ready plots, and
Scarab-formatted submission tables.

## Motivation for Bioconductor

Preclinical dose-response analysis sits between two well-served ecosystems
that leave a gap in the middle. On one side, generic curve-fitting
engines (`drc`, `nplr`, `dr4pl`) provide flexible non-linear regression
but expect the user to shape their own tidy data frame, pick a model
family, and hand-roll everything upstream and downstream of the fit —
plate parsing, control normalisation, outlier detection, batch iteration,
export. On the other side, large-scale pharmacogenomic frameworks
(`PharmacoGx`) treat drug response as one facet of curated multi-omic
screens (GDSC, CCLE, gCSI) via the `PharmacoSet` /
`TreatmentResponseExperiment` classes; they solve a different problem
than the day-to-day plate-reader output of a single wet lab.

`dosefitr` fills the gap: the user hands in a folder of BMG PHERAstar
Excel exports and receives a full analysis package back — normalised
ratio / viability tables, ROUT-flagged wells, 4PL fits with assay-aware
plausibility limits, quality-controlled `Curve_Quality` labels, per-plate
patchwork panels, cross-plate consistency checks, and a Scarab-formatted
submission table. No external pipeline, no shell scripts, no per-plate
manual data wrangling.

Inclusion in Bioconductor makes `dosefitr` discoverable to the
translational-pharmacology and chemical-biology communities that already
rely on Bioconductor for the omics half of their pipelines, and ensures
long-term maintenance through the Bioconductor release cycle.

## Comparison with related packages

Several CRAN and Bioconductor packages, and one non-R desktop tool,
address parts of the dose-response workflow:

- **`drc`** (CRAN) — the canonical generic non-linear regression
  toolkit for dose-response. Very flexible in model choice (log-logistic,
  Weibull, exponential decay, Brain-Cousens, ...), and the standard
  reference for methods research. `dosefitr` is not a replacement:
  it targets a narrower problem (kinase-binding NanoBRET and viability
  on 384-well PHERAstar plates) and layers plate parsing, control
  normalisation, ROUT outlier detection, batch iteration across plates,
  Curve_Quality classification, and Scarab-format export on top of the
  4PL / 3PL fits themselves.
- **`nplr`** (CRAN) — n-parameter logistic focused on IC50 estimation
  with confidence intervals. Single-curve API; the user shapes the
  data. Complementary to `dosefitr` for the fit itself but not for the
  surrounding pipeline (parsing, batching, QC, export).
- **`dr4pl`** (CRAN) — robust 4PL with convergence safeguards for
  pathological data (single high outliers, support problems, and
  non-logistic shapes that break `drc` and `nplr`) as documented in
  An et al. (R Journal, 2019). Single-curve API, complementary to
  `dosefitr` at the fit step; `dosefitr` uses its own `OptimModel`-based
  4PL / 3PL solver with assay-aware Bottom / Top / Hill-slope bounds.
- **`PharmacoGx`** (Bioconductor) — large-scale pharmacogenomic analysis
  built around the `PharmacoSet` and `TreatmentResponseExperiment` S4
  classes. Designed for cross-cell-line meta-analyses of published
  screens (GDSC, CCLE, gCSI), including drug-synergy detection
  (Loewe / Bliss / HSA / ZIP) from PharmacoGx 3.0 onwards. It is the
  right tool when the question is *how does this compound behave across
  a panel of curated cell lines and their molecular features?* —
  a different problem from *how do I turn today's PHERAstar output into
  a submission-ready IC50 table?*.
- **GraphPad Prism** (non-R desktop tool) — the community-standard
  interactive tool for benchtop dose-response analysis. Reproduces the
  4PL / 3PL fits and IC50 reporting that `dosefitr` provides, but is
  closed-source, licence-gated, and requires manual per-plate data
  wrangling. `dosefitr` reproduces its analytical outputs in a
  scriptable, batch-friendly, open-source R workflow.

`dosefitr`'s unique combination is: **BMG PHERAstar-aware parsing +
control-normalised ratio / viability tables + ROUT outlier detection +
assay-aware 4PL / 3PL fitting with Curve_Quality labels + Scarab-format
export**, all in one package with no external pipeline dependency.

## Vignette scope

This vignette introduces the package, walks a hero example on the
bundled NanoBRET plates, and points to the three companion vignettes
that cover the full workflows end-to-end.

- **NanoBRET pipeline** (dual-channel donor / acceptor, function by
  function): `vignette("dosefitr-nanobret")`.
- **Viability pipeline** (single-channel luminescence, e.g. CellTiter-Glo):
  `vignette("dosefitr-viability")`.
- **End-to-end protocol** (a working template mirroring a real analysis
  session, one topic per step): `vignette("dosefitr-protocol")`.

# Installation

`dosefitr` is planned for submission to Bioconductor once the
accompanying manuscript is published. Once accepted, install the release
version with:

```{r install-bioc, eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dosefitr")
```

The in-development version is on GitHub:

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("AKKDataAnalysis/dosefitr")
```

# Quick start

```{r setup}
library(dosefitr)
```

`dosefitr` ships two example NanoBRET plates and an info table under
`inst/extdata/`. We stage them into a working directory under
`tempdir()` to keep the vignette compliant with the "no writes outside
`tempdir()`" policy:

```{r stage}
extdata_dir <- system.file("extdata", package = "dosefitr")
work_dir    <- file.path(tempdir(), "dosefitr_intro")
dir.create(work_dir, showWarnings = FALSE)
invisible(file.copy(
    from      = list.files(extdata_dir, pattern = "^nanobret_",
                           full.names = TRUE),
    to        = work_dir,
    overwrite = TRUE
))
list.files(work_dir)
```

The hero example takes the two staged plates through the full pipeline:
BRET ratios, 4PL fits, per-compound summary, and one plot per compound.

```{r hero}
# 1. BRET ratios
ratio_res <- batch_ratio_analysis(
    directory        = work_dir,
    info_file        = "nanobret_info.xlsx",
    data_pattern     = "nanobret_plate_\\d+\\.xlsx$",
    control_0perc    = "1",
    control_100perc  = "24",
    selected_columns = 1:24,
    generate_reports = FALSE,
    output_dir       = file.path(tempdir(), "dosefitr_intro_ratio"),
    verbose          = FALSE
)

# 2. 4PL dose-response fits
drc_res <- batch_drc_analysis(
    batch_results    = ratio_res,
    model            = "4pl",
    normalize        = FALSE,
    output_dir       = file.path(tempdir(), "dosefitr_intro_drc"),
    generate_reports = FALSE,
    verbose          = FALSE
)

# 3. Per-plate summary table
sum_01 <- drc_res$drc_results$plate_01$drc_result$summary_table
sum_01[, c("Compound", "LogIC50", "HillSlope",
           "R_squared", "Curve_Quality")]
```

The `Curve_Quality` column carries per-compound labels (`Good curve`,
`Wide logIC50 CI range`, `Top too low`, `Bottom too high`, ...) that
distinguish clean fits from those with wide IC50 CIs, hill-slope
problems, or plateaux that do not bracket the response range. For the
bundled example plates almost every compound earns `Good curve`:

```{r quality}
table(sum_01$Curve_Quality)
```

For the full pipeline — plate layout, info table, ROUT outlier
detection, per-plate panel plots, cross-plate comparison, Scarab
export, and quality metrics — see:

- `vignette("dosefitr-nanobret", package = "dosefitr")` — BRET ratios,
  ROUT, 4PL fits, plotting, Scarab export.
- `vignette("dosefitr-viability", package = "dosefitr")` — single-channel
  viability normalisation, 4PL fits, Scarab export.

# Where to next

- `?batch_ratio_analysis` and `?batch_viability_analysis` — the two
  entry points, one per pipeline.
- `?batch_drc_analysis` — parameter reference for the shared fitting
  step, including the assay-aware plausibility limits and the full list
  of `Curve_Quality` labels.
- `?rout_outliers_batch` — ROUT outlier detection at the batch level.
- `?batch_save_all_drc_plots` — the workhorse batch plotter.
- `?scarab_table` and `?scarab_viability` — Scarab-format submission
  tables for NanoBRET and viability respectively.
- GitHub: <https://github.com/AKKDataAnalysis/dosefitr>

# Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
