---
title: "Introduction to dosefitr"
author: "Thiago Loreto Matos"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to dosefitr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

`dosefitr` provides an end-to-end dose-response analysis pipeline for two
assay types commonly used in drug discovery:

- **NanoBRET kinase binding assays** — reads raw BMG PHERAstar plate-reader
  exports, computes BRET ratios, and derives IC50 / pIC50 values for kinase
  inhibitors.
- **Cell viability assays** — reads luminescence or fluorescence plate-reader
  exports and fits dose-response curves to quantify compound cytotoxicity.

Both workflows share the same pipeline structure and downstream functions
(outlier removal, curve fitting, plotting, and Excel export), so switching
between assay types requires changing only one or two arguments.

### Pipeline overview

```
Raw Excel files
      │
      ▼
batch_ratio_analysis()        ← NanoBRET: reads plates, computes BRET ratios
batch_viability_analysis()    ← Viability: reads plates, normalises signal
      │
      ▼
rout_outliers_batch()         ← detects and removes outliers  [optional]
      │
      ▼
batch_drc_analysis()          ← fits 3PL or 4PL dose-response curves
      │
      ├──▶ batch_save_all_drc_plots()   ← one plot per compound
      ├──▶ plot_multiple_compounds()    ← overlay selected compounds
      ├──▶ compare_plates_drc()         ← compare compound across plates
      ├──▶ reshape_dr_table()           ← reshape summary for Excel
      ├──▶ save_multiple_sheets()       ← write multiple sheets to Excel
      └──▶ scarab_table()              ← Scarab-format export (NanoBRET)
```

---

## Installation

```{r install}
# Install from GitHub
remotes::install_github("AKKDataAnalysis/dosefitr")
```

```{r load}
library(dosefitr)
```

---

## Input files

### Directory layout

All files for a single experiment should live in the same folder:

```
experiment_folder/
├── info_tables.xlsx      ← metadata file (one sheet per plate)
├── raw_data_1.xlsx       ← plate-reader export for plate 1
├── raw_data_2.xlsx       ← plate-reader export for plate 2
└── raw_data_3.xlsx       ← plate-reader export for plate 3
```

Each sheet in `info_tables.xlsx` is matched to a raw data file by the
trailing number in the filename (e.g. sheet `Plate_1` → `raw_data_1.xlsx`).

### The metadata file (`info_tables.xlsx`)

This is the only file you need to create manually. It is an Excel workbook
with **one sheet per plate**. Each sheet must contain at least four columns
(column names can be anything; position matters):

| Position | Content | Example |
|---|---|---|
| 1 | Log~10~ inhibitor concentrations | `-9`, `-8.5`, `-8`, ... |
| 2 | Plate row letter(s) for this compound | `A`, `B`, `C`, ... |
| 3 | Construct / kinase name | `LRRK2`, `AAK1` |
| 4 | Compound / inhibitor name | `Compound_A`, `GSK123` |

A minimal example for two compounds tested at eight concentrations in
duplicate plate rows:

| log(inhibitor) | Plate_Row | Construct | Compound   |
|---|---|---|---|
| -9.0 | A | LRRK2 | Compound_A |
| -8.5 | B | LRRK2 | Compound_A |
| -8.0 | C | LRRK2 | Compound_A |
| -7.5 | D | LRRK2 | Compound_A |
| -7.0 | E | LRRK2 | Compound_A |
| -6.5 | F | LRRK2 | Compound_A |
| -6.0 | G | LRRK2 | Compound_A |
| -5.5 | H | LRRK2 | Compound_A |
| -9.0 | I | LRRK2 | Compound_B |
| -8.5 | J | LRRK2 | Compound_B |
| ...  | ... | ... | ... |

Columns 3 and 4 are combined as `Construct:Compound` (e.g. `LRRK2:Compound_A`)
to label all output columns, plots, and export tables throughout the pipeline.

---

## Workflow 1 — NanoBRET kinase binding assay

### Step 1: Process raw plates

`batch_ratio_analysis()` reads all PHERAstar Excel files in the working
directory, pairs them with the metadata file, computes BRET ratios
(acceptor / donor × 1000), and returns a named list with one entry per plate.

```{r nanobret-step1}
results <- batch_ratio_analysis(
  control_0perc       = 24,    # plate column for 0% control (e.g. DMSO)
  control_100perc     = 12,    # plate column for 100% control (e.g. staurosporine)
  output_dir          = "./qc_output",
  low_value_threshold = 1000,  # donor-channel values below this → NA
  verbose             = TRUE
)
```

Wells where the donor-channel signal falls below `low_value_threshold` are
set to `NA` before the ratio is calculated, excluding failed or empty wells.

**Using duplicate 100% control columns (v2 mode):**

When the 100% control is measured in two columns that should be averaged, or
when the 0% control is a fixed instrument background value, use
`function_version = "v2"`:

```{r nanobret-step1-v2}
results <- batch_ratio_analysis(
  control_0perc       = 16,         # fixed background value
  control_100perc     = c(12, 24),  # two columns averaged into one
  low_value_threshold = 3000,
  function_version    = "v2",
  verbose             = TRUE
)
```

### Step 2: Remove outliers (optional)

`rout_outliers_batch()` applies the ROUT method to each compound curve
independently. The `Q` parameter is the maximum acceptable false discovery
rate for outlier labelling (lower = more conservative).

```{r nanobret-step2}
results_clean <- rout_outliers_batch(results, Q = 0.01)

# Inspect flagged points
head(results_clean$outlier_summary)

# Visualise outlier-flagged curves
plot_outliers_batch_curves(results_clean)
```

`rout_outliers_batch()` overwrites `modified_ratio_table` in-place with the
cleaned data and preserves the original as `modified_ratio_table_original`.
Pass `results_clean` (not `results`) to `batch_drc_analysis()` to use the
cleaned data.

### Step 3: Fit dose-response curves

`batch_drc_analysis()` fits a three-parameter logistic (3PL) model to each
compound curve and returns IC50, pIC50, R², Hill slope, and confidence
intervals.

```{r nanobret-step3}
drc_results <- batch_drc_analysis(
  batch_results = results_clean,  # use results if you skipped Step 2
  normalize     = TRUE,
  assay_type    = "nanobret",
  output_dir    = "./drc_results",
  verbose       = TRUE
)
```

**3PL vs 4PL:**

- `model = "3pl"` (default): Hill slope fixed at −1. Recommended for standard
  sigmoidal inhibition curves with fewer free parameters.
- `model = "4pl"`: Hill slope freely estimated. Use when cooperative binding
  or mixed mechanisms are expected.

**Handling non-inhibitory curves:**

Flat curves (no meaningful response across the concentration range) always
have their IC50 and pIC50 reported as `"N/D"` (not determined). To also
suppress IC50 for activation curves (response increases with concentration):

```{r nanobret-step3-nd}
drc_results <- batch_drc_analysis(
  batch_results    = results_clean,
  normalize        = TRUE,
  nd_if_activation = TRUE   # activation curves → N/D
)
```

**Accessing results:**

```{r nanobret-step3-access}
# Final summary table for plate 1 (IC50, pIC50, R², Hill slope, ...)
summary_table <- drc_results$drc_results$plate_01$drc_result$final_summary_table

# Quality metrics table
quality_table <- drc_results$drc_results$plate_01$drc_result$curve_quality_table
```

### Step 4: Visualise results

**Save one plot per compound:**

```{r nanobret-step4-all}
batch_save_all_drc_plots(drc_results, verbose = TRUE)
```

**Overlay selected compounds on one plot:**

```{r nanobret-step4-overlay}
# By index
plot_multiple_compounds(drc_results,
  plate            = "plate_01",
  compound_indices = 1:6,
  color_palette    = "colorblind",
  y_limits         = c(0, 100),   # or NULL to auto-scale to the data
  y_axis_title     = "Normalized BRET ratio [%]",
  show_error_bars  = TRUE,
  save_plot        = TRUE
)

# By name or partial match
plot_multiple_compounds(drc_results, target_compound = "LRRK2")

# Additional appearance controls
plot_multiple_compounds(drc_results,
  compound_indices       = 1:6,
  x_limits               = c(-9, -5),   # log10 molar; NULL auto-scales
  x_limits_scale         = "log10",      # "log10", "molar", "uM", or "nM"
  x_axis_title           = NULL,         # NULL uses default Log10 Concentration [M]
  curve_linewidth        = 1,
  curve_alpha            = 0.7,
  show_ic50_lines        = TRUE,         # dashed vertical line at each IC50
  plot_title_size        = 16,
  axis_line_color        = "black",
  show_border            = FALSE,
  transparent_background = FALSE
)
```

**Compare the same compound across plates:**

```{r nanobret-step4-compare}
compare_plates_drc(
  drc_results,
  compare_by    = "compound",
  color_palette = "set1",
  y_limits      = c(0, 100),   # or NULL to auto-scale to the data
  y_axis_title  = "Normalized BRET ratio [%]",
  min_plates    = 2,
  # Additional appearance controls (same as plot_multiple_compounds)
  x_limits               = NULL,
  show_ic50_lines        = FALSE,
  axis_line_color        = "black",
  show_border            = FALSE,
  transparent_background = FALSE
)
```

### Step 5: Export results

**Reshape the summary table and save to Excel:**

```{r nanobret-step5-reshape}
final_table <- drc_results$drc_results$plate_01$drc_result$final_summary_table

excel_table <- reshape_dr_table(
  final_table,
  decimal_comma = FALSE,
  output_file   = "results_plate01.xlsx"
)
```

**Save multiple data frames to one workbook:**

```{r nanobret-step5-multi}
original_table <- results_clean$plate_01$result$modified_ratio_table_original

save_multiple_sheets(
  "final_data.xlsx",
  original_table,
  excel_table,
  decimal_comma  = TRUE,
  decimal_places = 3,
  round_sheets   = 1:2
)
```

**Generate a Scarab-format export table (NanoBRET):**

`scarab_table()` produces a 76-row Scarab-format data frame combining
experimental metadata, curve parameters, QC metrics, and raw BRET values.
The output is saved as an `.xlsx` file with two sheets: `Scarab_Table`
(fields as rows, compounds as columns) and `Scarab_Table_Transposed`
(compounds as rows — easier for filtering in Excel).

```{r nanobret-step5-scarab}
scarab_table(
  results_list              = results,
  drc_results_list          = drc_results,
  plate_name                = "plate_01",
  date                      = "260323",
  experimenter_abbrev       = "TL",
  nLuc_orientation          = "C",       # "N" or "C" terminus
  tracer                    = "Tracer K10",
  tracer_kd_app             = -7.5,
  tracer_concentration_used = -7.5,
  decimal_separator         = ",",       # use "." for English format
  eubscarab_ready           = "No"
)
```

Per-kinase parameters (`nLuc_orientation`, `tracer`, `tracer_kd_app`,
`tracer_concentration_used`) accept a single value (applied to all kinases),
a named vector (keyed by kinase name), or an unnamed vector (applied in order
of unique kinases on the plate):

```{r nanobret-step5-scarab-multi}
scarab_table(
  results_list     = results,
  drc_results_list = drc_results,
  plate_name       = "plate_01",
  date             = "260323",
  experimenter_abbrev = "TL",
  nLuc_orientation = c(LRRK2 = "C", AAK1 = "N"),
  tracer           = c(LRRK2 = "Tracer K10", AAK1 = "Tracer 236"),
  tracer_kd_app    = c(LRRK2 = -7.5, AAK1 = -8.0),
  tracer_concentration_used = c(LRRK2 = -7.5, AAK1 = -8.0),
  decimal_separator = ","
)
```

---

## Workflow 2 — Cell viability assay

The viability workflow is structurally identical to the NanoBRET workflow.
Replace `batch_ratio_analysis()` with `batch_viability_analysis()` in Step 1
and pass `assay_type = "viability"` to `batch_drc_analysis()` in Step 3.
All other functions work without modification.

### Step 1: Process raw plates

```{r viability-step1}
via_results <- batch_viability_analysis(
  directory       = "data/viability/",
  control_0perc   = 1,    # column for background / vehicle control
  control_100perc = 24,   # column for untreated cells (maximum signal)
  output_dir      = "./qc_output",
  verbose         = TRUE
)
```

To exclude outer control columns from the compound data:

```{r viability-step1-cols}
via_results <- batch_viability_analysis(
  control_0perc    = 13,
  control_100perc  = 12,
  selected_columns = 2:23   # exclude columns 1 and 24
)
```

Per-plate quality metrics (CV%, Z-prime, signal-to-background) are computed
automatically and saved in the `Quality_Metrics` sheet of each per-plate
Excel file.

### Step 2: Remove outliers (optional)

Use `keep_cytotoxic = TRUE` to retain data points that resemble outliers but
are part of a genuine cytotoxic response (sharp drop at high concentrations):

```{r viability-step2}
results_clean <- rout_outliers_batch(via_results,
  Q              = 0.01,
  keep_cytotoxic = TRUE
)
```

### Step 3: Fit dose-response curves

```{r viability-step3}
drc_results <- batch_drc_analysis(
  batch_results = results_clean,  # use via_results if you skipped Step 2
  normalize     = TRUE,
  assay_type    = "viability",    # applies viability-appropriate parameter limits
  output_dir    = "./drc_results",
  verbose       = TRUE
)
```

### Step 4: Visualise and export

All plotting and export functions work identically to the NanoBRET workflow.
Pass a custom y-axis label where appropriate:

```{r viability-step4}
batch_save_all_drc_plots(drc_results,
  verbose      = TRUE,
  y_axis_title = "Cell Viability (%)",
  y_limits     = NULL   # NULL auto-scales each plot; use c(0, 100) for a fixed scale
)

plot_multiple_compounds(drc_results,
  compound_indices = 1:6,
  color_palette    = "colorblind",
  y_limits         = c(0, 100),   # or NULL to auto-scale to the data
  y_axis_title     = "Cell Viability (%)",
  save_plot        = TRUE
)

# Additional appearance controls (same as NanoBRET workflow)
plot_multiple_compounds(drc_results,
  compound_indices       = 1:6,
  x_limits               = c(-9, -5),   # log10 molar; NULL auto-scales
  x_limits_scale         = "log10",      # "log10", "molar", "uM", or "nM"
  x_axis_title           = NULL,
  curve_linewidth        = 1,
  curve_alpha            = 0.7,
  show_ic50_lines        = TRUE,
  plot_title_size        = 16,
  axis_line_color        = "black",
  show_border            = FALSE,
  transparent_background = FALSE
)
```

---

## Merging replicate plates

When the same experiment is run across multiple plates (biological
replicates), `merge_plate_replicates()` combines their data tables into a
single entry that feeds directly into all downstream functions.

```{r merge}
# Merge all plates
merged <- merge_plate_replicates(results)

# Merge a subset, keeping others separate
merged <- merge_plate_replicates(results,
  plates      = c("plate_01", "plate_02"),
  merged_name = "rep1_rep2"
)

# Save the Excel report to a custom directory (default: drc_quality/ in working dir)
merged <- merge_plate_replicates(results, output_dir = "./my_reports")

# Feed into the DRC step as usual
drc_results <- batch_drc_analysis(
  batch_results = merged,
  normalize     = TRUE,
  output_dir    = "./drc_results"
)
```

Replicate columns are renumbered sequentially across plates
(`LRRK2:Compound_A`, `LRRK2:Compound_A.2` from plate 1 become
`LRRK2:Compound_A.3`, `LRRK2:Compound_A.4` from plate 2, and so on).
Compounds that appear in only some plates are included with however many
replicates they have — no NA-padding is added.

---

## Complete example — NanoBRET

```{r full-nanobret}
library(dosefitr)

setwd("path/to/experiment")

# 1. Process raw plates
results <- batch_ratio_analysis(
  control_0perc       = 24,
  control_100perc     = 12,
  output_dir          = "./qc_output",
  low_value_threshold = 1000,
  verbose             = TRUE
)

# 2. Remove outliers
results_clean <- rout_outliers_batch(results, Q = 0.01)
plot_outliers_batch_curves(results_clean)

# 3. Fit dose-response curves
drc_results <- batch_drc_analysis(
  batch_results = results_clean,
  normalize     = TRUE,
  output_dir    = "./drc_results",
  verbose       = TRUE
)

# 4. Save all individual plots
batch_save_all_drc_plots(drc_results, verbose = TRUE)

# 5. Overlay selected compounds
plot_multiple_compounds(drc_results,
  plate            = "plate_01",
  compound_indices = 1:6,
  color_palette    = "colorblind",
  y_limits         = c(0, 100),   # or NULL to auto-scale
  y_axis_title     = "Normalized BRET ratio [%]",
  save_plot        = TRUE
)

# 6. Compare the same compound across plates
compare_plates_drc(drc_results, compare_by = "compound")

# 7. Export Scarab table
scarab_table(
  results_list              = results,
  drc_results_list          = drc_results,
  plate_name                = "plate_01",
  date                      = "260323",
  experimenter_abbrev       = "TL",
  nLuc_orientation          = "C",
  tracer                    = "Tracer K10",
  tracer_kd_app             = -7.5,
  tracer_concentration_used = -7.5,
  decimal_separator         = ",",
  eubscarab_ready           = "No"
)
```

---

## Complete example — Cell viability

```{r full-viability}
library(dosefitr)

setwd("path/to/experiment")

# 1. Process raw plates
via_results <- batch_viability_analysis(
  control_0perc    = 13,
  control_100perc  = 12,
  selected_columns = 2:23,
  verbose          = TRUE
)

# 2. Remove outliers (keep cytotoxic points)
results_clean <- rout_outliers_batch(via_results,
  Q              = 0.01,
  keep_cytotoxic = TRUE
)
plot_outliers_batch_curves(results_clean)

# 3. Fit dose-response curves
drc_results <- batch_drc_analysis(
  batch_results = results_clean,
  normalize     = TRUE,
  assay_type    = "viability",
  output_dir    = "./drc_results",
  verbose       = TRUE
)

# 4. Save all individual plots
batch_save_all_drc_plots(drc_results,
  verbose      = TRUE,
  y_axis_title = "Cell Viability (%)"
)

# 5. Overlay selected compounds
plot_multiple_compounds(drc_results,
  compound_indices = 1:6,
  color_palette    = "colorblind",
  y_limits         = c(0, 100),   # or NULL to auto-scale
  y_axis_title     = "Cell Viability (%)",
  save_plot        = TRUE
)

# 6. Export summary table
final_table <- drc_results$drc_results$plate_01$drc_result$final_summary_table
excel_table <- reshape_dr_table(final_table,
  decimal_comma = FALSE,
  output_file   = "results.xlsx"
)

# 7. Save combined workbook
original_table <- results_clean$plate_01$result$modified_ratio_table_original
save_multiple_sheets("final_data.xlsx",
  original_table, excel_table,
  decimal_comma  = TRUE,
  decimal_places = 3,
  round_sheets   = 1:2
)
```

---

## Function reference

| Function | Description |
|---|---|
| `batch_ratio_analysis()` | Read raw NanoBRET plates, compute BRET ratios |
| `batch_viability_analysis()` | Read raw viability plates, normalise signal |
| `merge_plate_replicates()` | Merge replicate plates into a single combined result |
| `rout_outliers_batch()` | ROUT outlier detection across all plates |
| `plot_outliers_batch_curves()` | Visualise outlier-flagged curves |
| `batch_drc_analysis()` | Fit 3PL or 4PL dose-response curves across all plates |
| `fit_drc_3pl()` | Fit a single plate with a 3-parameter logistic model |
| `fit_drc_4pl()` | Fit a single plate with a 4-parameter logistic model |
| `batch_save_all_drc_plots()` | Save one plot per compound across all plates |
| `plot_multiple_compounds()` | Overlay selected compounds on one plot |
| `compare_plates_drc()` | Compare the same compound across plates |
| `reshape_dr_table()` | Reshape summary table for Excel export |
| `save_multiple_sheets()` | Save multiple data frames to one Excel workbook |
| `scarab_table()` | Generate Scarab-format export table (NanoBRET) |
