---
title: "NanoBRET pipeline: dose-response analysis end-to-end"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{NanoBRET pipeline: dose-response analysis end-to-end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

## Overview

This vignette walks through the **NanoBRET** dose-response pipeline of
**dosefitr**, end-to-end, on bundled example data. Every step runs at
build time on two small plates included with the package, so you can
copy any chunk straight into your own R session.

The eight stages of the pipeline:

1.  Setup and stage the bundled plates into a working directory.
2.  Inspect the plate layout and the info table.
3.  Compute BRET ratios from the dual-emission raw data
    (`batch_ratio_analysis()`).
4.  Detect outliers with the ROUT test
    (`rout_outliers_batch()` + `plot_outliers_curves()`).
5.  Fit 4PL dose-response curves
    (`batch_drc_analysis()`).
6.  Visualise individual and overlay plots
    (`batch_save_all_drc_plots()`, `plot_multiple_compounds()`,
    `compare_plates_drc()`).
7.  Export Scarab-format tables and multi-sheet workbooks
    (`scarab_table()`, `save_multiple_sheets()`).
8.  Read the quality metrics.

## 1. Setup

```{r libs}
library(dosefitr)
```

The package ships with two example PHERAstar plates and their
accompanying info table. They are located in `inst/extdata/` and are
loaded with `system.file()`:

```{r locate-data}
extdata_dir <- system.file("extdata", package = "dosefitr")

list.files(extdata_dir, pattern = "^nanobret_")
```

`batch_ratio_analysis()` scans a directory for plate files matching a
regular expression, so we stage the relevant files into a working
directory under `tempdir()`. Writing to `tempdir()` keeps the vignette
compliant with CRAN policy (no writes outside the temporary directory):

```{r stage}
work_dir <- file.path(tempdir(), "dosefitr_nanobret")
dir.create(work_dir, showWarnings = FALSE)

invisible(file.copy(
  from      = list.files(extdata_dir, pattern = "^nanobret_",
                         full.names = TRUE),
  to        = work_dir,
  overwrite = TRUE
))

list.files(work_dir)
```

## 2. Plate layout and info table

Each plate is a 384-well PHERAstar export with two emission channels
(donor 450-80 nm, acceptor 610-80 nm). The first 12 rows of the
worksheet carry instrument metadata; the data is laid out in 16 rows
(`A`-`P`) by 24 columns:

-   **Column 1** -- DMSO (vehicle) control, the "0 % inhibition" baseline.
-   **Columns 2-12** -- 11 inhibitor concentrations, technical replicate 1.
-   **Columns 13-23** -- the same 11 concentrations, technical replicate 2.
-   **Column 24** -- saturated competitor, the "100 % inhibition" floor.

The mapping from row to compound is in `nanobret_info.xlsx`, one sheet
per plate. The first column lists log10-concentrations (with `NA` for
control rows), and the remaining columns identify each plate row:

```{r show-info}
info_path <- file.path(work_dir, "nanobret_info.xlsx")

info_plate_01 <- openxlsx::read.xlsx(info_path, sheet = "plate_01")

head(info_plate_01, 8)
```

The example data covers 16 compounds per plate. Plate 01 has three
targets (`KinaseA`, `KinaseB`, `KinaseC`); plate 02 has one
(`KinaseD`) plus four "bridging" rows that repeat compounds from
plate 01 to demonstrate the inter-plate comparison step.

## 3. BRET ratio computation

`batch_ratio_analysis()` reads both plates, extracts the donor and
acceptor matrices, computes the BRET ratio per well as
`(acceptor / donor) * 1000`, and normalises each row by its controls.

```{r ratio}
ratio_out <- file.path(tempdir(), "dosefitr_nanobret_ratio")
dir.create(ratio_out, showWarnings = FALSE)

ratio_res <- batch_ratio_analysis(
  directory        = work_dir,
  info_file        = "nanobret_info.xlsx",
  data_pattern     = "nanobret_plate_\\d+\\.xlsx$",
  control_0perc    = "1",
  control_100perc  = "24",
  selected_columns = 1:24,
  generate_reports = FALSE,
  output_dir       = ratio_out,
  verbose          = FALSE
)

names(ratio_res)
```

The returned object is a named list, one entry per plate. The actual
ratio tables live under `$result`:

```{r ratio-inspect}
mrt <- ratio_res$plate_01$result$modified_ratio_table

dim(mrt)

head(mrt[, 1:6], 8)
```

A well-behaved inhibition curve runs from a high BRET ratio at the
lowest concentrations (tracer bound) down to a low BRET ratio at the
highest concentrations (tracer displaced). Looking at one compound:

```{r ratio-direction}
data.frame(
  log_conc = mrt[, 1],
  Cpd1_rep1 = round(mrt[, "KinaseA:Cpd1"],   1),
  Cpd1_rep2 = round(mrt[, "KinaseA:Cpd1.2"], 1)
)
```

The control rows (first and last, with `NA` log-concentration) carry
the 0 % and 100 % inhibition control means.

## 4. Outlier detection (ROUT)

`rout_outliers_batch()` runs the ROUT test
(Motulsky & Brown 2006, BMC Bioinformatics) on every fitted curve.
Wells that diverge by more than the ROUT Q-threshold (default 1 %)
**and** disagree with their replicate are flagged:

```{r rout}
rout_res <- rout_outliers_batch(
  batch_results = ratio_res,
  Q             = 0.01,
  n_param       = 4L,
  direction     = "inhibition",
  verbose       = FALSE
)

# Summary of detection
nrow(rout_res$outlier_summary)

table(rout_res$outlier_summary$plate)
```

`outlier_summary` has one row per flagged well, with the fitted value,
the observed value, the standardised residual, and the dynamic range:

```{r rout-summary}
head(rout_res$outlier_summary, 6)
```

`plot_outliers_curves()` overlays the flagged wells on the fitted
curves, so you can sanity-check the call:

```{r rout-plot, fig.height = 4}
plot_outliers_curves(
  rout_output   = rout_res$plate_01,
  subplot_title = "compound",
  ncol          = 4L
)
```

For most clean plates, ROUT flags single outliers in only a handful of
compounds; large clusters of flags usually indicate a misfit rather
than a true outlier.

## 5. Dose-response curve fitting

`batch_drc_analysis()` fits each compound's BRET ratio with a 4PL
model. By default it auto-detects the assay type from the ratio table
shape (here, `nanobret`) and applies assay-aware plausibility limits
on the Bottom, Top, and Hill parameters.

```{r drc}
drc_out <- file.path(tempdir(), "dosefitr_nanobret_drc")
dir.create(drc_out, showWarnings = FALSE)

drc_res <- batch_drc_analysis(
  batch_results    = ratio_res,
  model            = "4pl",
  normalize        = FALSE,
  output_dir       = drc_out,
  generate_reports = FALSE,
  verbose          = FALSE
)

drc_res$drc_results$plate_01$drc_result$successful_fits
```

The per-plate summary table carries the fitted parameters, 95 %
confidence intervals, R^2, and a quality label:

```{r drc-summary}
sum_01 <- drc_res$drc_results$plate_01$drc_result$summary_table

sum_01[, c("Compound", "LogIC50", "HillSlope",
           "R_squared", "Curve_Quality")]
```

Quality labels distinguish clean fits from those with wide IC50
confidence intervals, hill-slope problems, or other issues. For this
clean example data set, almost every compound earns "Good curve":

```{r drc-quality}
table(sum_01$Curve_Quality)
```

## 6. Visualisation

### 6.1 Save one plot per compound (and a combined panel per plate)

`batch_save_all_drc_plots()` is the workhorse plotting function for a
batch result. It calls `plot_dose_response()` on every successful fit,
writes one PNG per compound, and assembles the per-plate compounds
into a single panel image with **patchwork**:

```{r batch-save-plots}
plots_dir <- file.path(tempdir(), "dosefitr_nanobret_plots")
dir.create(plots_dir, showWarnings = FALSE)

results2 <- batch_save_all_drc_plots(
  batch_drc_results = drc_res,
  output_dir        = plots_dir,
  verbose           = FALSE,
  dpi               = 120,            # lighter for the vignette
  panel_width_per_col  = 4,
  panel_height_per_row = 4,
  panel_spacing        = 1.2
)

results2$successes
basename(results2$panel_files)
```

Each plate gets its own subfolder under `output_dir` with the
individual compound PNGs and a single `<plate>_panel.png` overview.
Here is the assembled panel for plate 01:

```{r show-panel, echo = FALSE, out.width = "100%"}
knitr::include_graphics(results2$panel_files[1])
```

### 6.2 Multiple compounds overlaid

`plot_multiple_compounds()` overlays several curves to highlight
differences in potency or shape:

```{r plot-overlay}
plot_multiple_compounds(
  results          = drc_res,
  compound_indices = 1:4,
  plate            = "plate_01",
  y_limits         = NULL,
  show_error_bars  = TRUE,
  verbose          = FALSE
)
```

### 6.3 The same compound across plates

`compare_plates_drc()` makes one figure per construct-compound that
appears on more than one plate, useful for cross-plate consistency
QC. The bundled `plate_02` repeats four of `plate_01`'s compounds
under the same construct, so four cross-plate plots are produced:

```{r compare-plates, fig.show = "hide"}
compare_out <- file.path(tempdir(), "dosefitr_nanobret_compare")
dir.create(compare_out, showWarnings = FALSE)

cmp_res <- compare_plates_drc(
  batch_drc_result = drc_res,
  compare_by       = "compound",
  output_dir       = compare_out,
  min_plates       = 2,
  verbose          = FALSE
)

list.files(compare_out, pattern = "\\.png$")
```

`compare_plates_drc()` returns plot objects in `$plots` and writes
each one to disk; you can inspect them with any image viewer or load
them inline.

## 7. Export

### 7.1 Scarab-format submission tables

`scarab_table()` produces a wide submission table in the SGC Scarab
format, one column per compound:

```{r scarab}
sc_01 <- scarab_table(
  results_list      = ratio_res,
  drc_results_list  = drc_res,
  plate_name        = "plate_01",
  date              = format(Sys.Date(), "%y%m%d"),
  experimenter_abbrev = "DEMO",
  save              = FALSE
)

dim(sc_01)

head(sc_01[, 1:5], 6)
```

### 7.2 Multi-sheet Excel workbook

`save_multiple_sheets()` writes any number of data frames to a single
Excel file, one sheet per argument. Here we export both the ratio
table and the fit summary for plate 01:

```{r save-multi}
xlsx_out <- file.path(tempdir(), "dosefitr_nanobret_export.xlsx")

save_multiple_sheets(
  file_name     = xlsx_out,
  Ratios        = mrt,
  Fit_Summary   = sum_01,
  decimal_comma = FALSE
)

file.exists(xlsx_out)
```

## 8. Reading the quality metrics

The result of `batch_ratio_analysis()` includes a per-plate quality
block printed at run time. The key numbers to watch:

-   **S/B (signal-to-background)** -- ratio of mean 0 % control wells
    to mean 100 % control wells in the acceptor channel. Values
    above 5 indicate a usable dynamic range.

-   **Z-prime** -- a unitless robustness statistic combining S/B with
    control variability. Values above 0.5 are usable, above 0.7 are
    excellent.

-   **CV %** -- coefficient of variation between technical
    replicates. Anything under 10 % is well-behaved.

For a complete description of each fit-level quality label
(`Good curve`, `Wide logIC50 CI range`, `Top too low`, etc.) see
the help page of `batch_drc_analysis()`.

## See also

-   `vignette("dosefitr-viability", package = "dosefitr")` for the
    viability/proliferation pipeline using the same plate layout.
-   `?batch_ratio_analysis`, `?batch_drc_analysis`,
    `?rout_outliers_batch` for parameter-level documentation.
