---
title: "Viability pipeline: dose-response analysis end-to-end"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Viability pipeline: dose-response analysis end-to-end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  eval      = TRUE,
  message   = FALSE,
  warning   = FALSE
)
```

## Overview

This vignette walks through the **viability / proliferation**
dose-response pipeline of **dosefitr**, end-to-end, on bundled
example data. The pipeline is single-channel (CellTiter-Glo-style
luminescence), in contrast to the dual-channel NanoBRET pipeline
covered in `vignette("dosefitr-nanobret")`. The plate layout is the
same -- this matters because the two pipelines compose: one set of
compounds, one set of plate maps, two readouts.

The seven stages of the pipeline:

1.  Setup and stage the bundled plates into a working directory.
2.  Inspect the plate layout and the info table.
3.  Process the raw counts into normalised viability tables
    (`batch_viability_analysis()`).
4.  Fit 4PL dose-response curves
    (`batch_drc_analysis()` with `normalize = TRUE`).
5.  Visualise individual and overlay plots
    (`batch_save_all_drc_plots()`, `plot_multiple_compounds()`,
    `compare_plates_drc()`).
6.  Export Scarab-format tables
    (`scarab_viability()`).
7.  Read the quality metrics.

A short note at the end (Section 8) explains how to use the v2
processor when your info table follows the alternative
PHERAstar-with-merged-controls format.

## 1. Setup

```{r libs}
library(dosefitr)
```

The package ships with two viability plates and a single info table
covering them both. They are located in `inst/extdata/` and are
loaded with `system.file()`:

```{r locate-data}
extdata_dir <- system.file("extdata", package = "dosefitr")

list.files(extdata_dir, pattern = "^viability_")
```

`batch_viability_analysis()` scans a directory for plate files
matching a regular expression, so we stage the relevant files into a
working directory under `tempdir()`. This keeps the vignette CRAN-
compliant (no writes outside `tempdir()`):

```{r stage}
work_dir <- file.path(tempdir(), "dosefitr_viability")
dir.create(work_dir, showWarnings = FALSE)

invisible(file.copy(
  from      = list.files(extdata_dir, pattern = "^viability_",
                         full.names = TRUE),
  to        = work_dir,
  overwrite = TRUE
))

list.files(work_dir)
```

## 2. Plate layout and info table

Each plate is a 384-well PHERAstar luminescence export with a single
emission channel. The first 12 rows of the worksheet are instrument
metadata; the data block follows at row 13. Each of the 16 rows
(`A`-`P`) holds one compound:

-   **Column 1** -- DMSO (vehicle) control, the "100 % viability" baseline.
-   **Columns 2-12** -- 11 inhibitor concentrations, technical replicate 1.
-   **Columns 13-23** -- the same 11 concentrations, technical replicate 2.
-   **Column 24** -- compound-saturated wells, the "0 % viability" floor.

`viability_info.xlsx` carries one sheet per plate. The first column
holds log10-concentrations (with `NA` for the control rows), the
remaining columns identify each plate row:

```{r show-info}
info_path <- file.path(work_dir, "viability_info.xlsx")

info_plate_01 <- openxlsx::read.xlsx(info_path, sheet = "plate_01")

head(info_plate_01, 8)
```

The example data covers 16 compounds per plate. Plate 01 mixes three
constructs (`KinaseA`, `KinaseB`, `KinaseC`); plate 02 carries one
construct (`KinaseD`) plus four bridging rows that repeat plate 01
compounds to demonstrate the inter-plate consistency QC.

## 3. Viability table processing

`batch_viability_analysis()` reads both plates, normalises each row
to its DMSO control, applies the 0 % floor at the saturated-compound
control, and produces percent-viability tables ready for fitting.
For the bundled v1-format data we use `version = "v1"`:

```{r viability}
via_out <- file.path(tempdir(), "dosefitr_viability_proc")
dir.create(via_out, showWarnings = FALSE)

via_res <- batch_viability_analysis(
  directory        = work_dir,
  info_file        = "viability_info.xlsx",
  data_pattern     = "viability_plate_\\d+\\.xlsx$",
  control_0perc    = 1,
  control_100perc  = 24,
  selected_columns = 1:24,
  generate_reports = FALSE,
  output_dir       = via_out,
  verbose          = FALSE,
  version          = "v1"
)

names(via_res)
```

Note that viability accepts **integer** control column indices,
unlike NanoBRET which uses character indices.

The returned object mirrors the NanoBRET pipeline: one entry per
plate, with the processed table under `$result$modified_ratio_table`:

```{r viability-inspect}
mrt <- via_res$plate_01$result$modified_ratio_table

dim(mrt)

head(mrt[, 1:6], 8)
```

A well-behaved cytotoxicity curve runs from high signal at the lowest
concentrations (cells healthy) down to low signal at the highest
concentrations (cells killed):

```{r viability-direction}
data.frame(
  log_conc  = mrt[, 1],
  Cpd1_rep1 = round(mrt[, "KinaseA:Cpd1"],   0),
  Cpd1_rep2 = round(mrt[, "KinaseA:Cpd1.2"], 0)
)
```

The first and last rows (with `NA` log-concentration) carry the
0 % and 100 % control row averages.

## 4. Dose-response curve fitting

`batch_drc_analysis()` fits each compound with a 4PL model. For
viability data the standard call passes `normalize = TRUE`, which
rescales the raw counts to percent and clamps the fits to viability-
plausible Bottom and Top windows.

```{r drc}
drc_out <- file.path(tempdir(), "dosefitr_viability_drc")
dir.create(drc_out, showWarnings = FALSE)

drc_res <- batch_drc_analysis(
  batch_results    = via_res,
  model            = "4pl",
  normalize        = TRUE,
  output_dir       = drc_out,
  generate_reports = FALSE,
  verbose          = FALSE
)

drc_res$drc_results$plate_01$drc_result$successful_fits
```

The per-plate summary table reports each compound's fitted Bottom
and Top (in percent), LogIC50, Hill slope, R^2, and a quality label:

```{r drc-summary}
sum_01 <- drc_res$drc_results$plate_01$drc_result$summary_table

sum_01[, c("Compound", "LogIC50", "HillSlope",
           "R_squared", "Curve_Quality")]
```

For this clean bundled data, most compounds earn a `Good curve`
label:

```{r drc-quality}
table(sum_01$Curve_Quality)
```

## 5. Visualisation

### 5.1 Save one plot per compound (and a combined panel per plate)

`batch_save_all_drc_plots()` is the workhorse plotting function for a
batch result. It calls `plot_dose_response()` on every successful fit,
writes one PNG per compound, and assembles the per-plate compounds
into a single panel image with **patchwork**:

```{r batch-save-plots}
plots_dir <- file.path(tempdir(), "dosefitr_viability_plots")
dir.create(plots_dir, showWarnings = FALSE)

results2 <- batch_save_all_drc_plots(
  batch_drc_results = drc_res,
  output_dir        = plots_dir,
  verbose           = FALSE,
  dpi               = 120,            # lighter for the vignette
  panel_width_per_col  = 4,
  panel_height_per_row = 4,
  panel_spacing        = 1.2
)

results2$successes
basename(results2$panel_files)
```

Each plate gets its own subfolder under `output_dir` with the
individual compound PNGs and a single `<plate>_panel.png` overview.
Here is the assembled panel for plate 01:

```{r show-panel, echo = FALSE, out.width = "100%"}
knitr::include_graphics(results2$panel_files[1])
```

### 5.2 Multiple compounds overlaid

`plot_multiple_compounds()` is the canonical way to compare several
compounds in one figure, ranked by potency:

```{r plot-overlay}
plot_multiple_compounds(
  results          = drc_res,
  compound_indices = 1:4,
  plate            = "plate_01",
  show_error_bars  = TRUE,
  verbose          = FALSE
)
```

### 5.3 The same compound across plates

`compare_plates_drc()` makes one figure per construct-compound pair
that appears on more than one plate -- the inter-plate consistency
check. The bundled `plate_02` repeats four of `plate_01`'s compounds
under the same construct, so four cross-plate plots are produced:

```{r compare-plates, fig.show = "hide"}
compare_out <- file.path(tempdir(), "dosefitr_viability_compare")
dir.create(compare_out, showWarnings = FALSE)

cmp_res <- compare_plates_drc(
  batch_drc_result = drc_res,
  compare_by       = "compound",
  output_dir       = compare_out,
  min_plates       = 2,
  verbose          = FALSE
)

list.files(compare_out, pattern = "\\.png$")
```

The returned object's `$plots` list has one ggplot per pairing.

## 6. Export

`scarab_viability()` produces a wide submission table in the SGC
Scarab format for viability assays:

```{r scarab}
sc_01 <- scarab_viability(
  results_list      = via_res,
  drc_results_list  = drc_res,
  plate_name        = "plate_01",
  date              = format(Sys.Date(), "%y%m%d"),
  experimenter_abbrev = "DEMO",
  save              = FALSE
)

dim(sc_01)

head(sc_01[, 1:5], 6)
```

For a generic multi-sheet workbook, `save_multiple_sheets()` writes
any number of data frames to one Excel file, one sheet per argument:

```{r save-multi}
xlsx_out <- file.path(tempdir(), "dosefitr_viability_export.xlsx")

save_multiple_sheets(
  file_name     = xlsx_out,
  Viability     = mrt,
  Fit_Summary   = sum_01,
  decimal_comma = FALSE
)

file.exists(xlsx_out)
```

## 7. Reading the quality metrics

The output of `batch_viability_analysis()` includes per-plate quality
metrics, printed at run time when `verbose = TRUE`. Key numbers:

-   **S/B** -- ratio of mean 100 % control wells (DMSO) to mean 0 %
    control wells (compound-saturated). Above 5 is usable.

-   **Z-prime** -- robustness statistic combining S/B with control
    variability. Above 0.5 is usable, above 0.7 is excellent.

-   **CV %** -- coefficient of variation between technical replicates.
    Under 10 % is well-behaved.

The fit-level `Curve_Quality` column carries one of:

-   `Good curve` -- everything passed.
-   `Wide logIC50 CI range` -- the IC50 95 % CI is wider than the
    test concentration range; the IC50 is at or beyond a tested
    concentration.
-   `Top too low` / `Bottom too high` -- the fitted plateaux do not
    bracket the response range.

See `?batch_drc_analysis` for the complete list and thresholds.

## 8. Note on the v2 processor

For info tables that follow the alternative PHERAstar layout (with
merged control columns in a single sheet), pass `version = "v2"`:

```{r v2-note, eval = FALSE}
via_res_v2 <- batch_viability_analysis(
  directory        = work_dir,
  info_file        = "viability_info.xlsx",
  data_pattern     = "viability_plate_\\d+\\.xlsx$",
  control_0perc    = 1,
  control_100perc  = 24,
  selected_columns = 1:24,
  generate_reports = FALSE,
  output_dir       = via_out,
  version          = "v2"
)
```

The two processors share the downstream pipeline: the same
`batch_drc_analysis()` call, the same plot functions, the same
`scarab_viability()` export. The only structural difference is the
shape of the info table.

## See also

-   `vignette("dosefitr-nanobret", package = "dosefitr")` for the
    BRET-ratio pipeline using the same plate layout.
-   `?batch_viability_analysis`, `?batch_drc_analysis`,
    `?scarab_viability` for parameter-level documentation.
